\chapter*{Declaration}
\thispagestyle{empty}

This thesis is the result of my own work and includes nothing which is the outcome of work done in collaboration except as declared in the preface and specified in the text. It is not substantially the same as any work that has already been submitted, or is being concurrently submitted, for any degree, diploma or other qualification at the University of Cambridge or any other University or similar institution except as declared in the preface and specified in the text. It does not exceed the prescribed word limit for the relevant Degree Committee.

The research described in Chapter \ref{chap:jtd_tismir} was completed with assistance from undergraduate student interns Joshua Schlichting and Tessa Pastor, who I was responsible for supervising alongside Peter Harrison. Joshua and Tessa assisted in identifying recordings that met \replaced[id=HC]{the}{an} inclusion criteria defined \replaced[id=HC]{in }{by myself (}Section \ref{sec:jtd_inclusion_criteria}\deleted[id=HC]{)}. They each also contributed three of the annotation files used for evaluating the automated system (Section \ref{sec:jtd_validation}).

An early version of the methods described in Section \ref{sec:rsi_handcrafted_methods} were used in an undergraduate dissertation submitted by Reuben Bance to the Faculty of Music, University of Cambridge, in Summer 2024. I was responsible for supervising Reuben's work, alongside Peter Harrison, and developed these methods in collaboration with him. These methods were subsequently refined for inclusion in Chapter \ref{chap:xai_rsi} by myself, and I am otherwise responsible for the entirety of the modelling and analysis conducted in this chapter. 

The experiments described in Section \ref{sec:mp_methods} received ethical approval from the Ethics Review Committee at the Faculty of Music, University of Cambridge (reference number: 21-22-163). The experiment described in Section \ref{sec:gen_subjective_evaluation} received approval from the same committee (reference number: 2025-015).

% Leaving some space for the signature:
\vspace{15mm}

