\chapter*{Publications}

% Fake bibliography hanging paragraphs
\newcommand{\bibitemhang}[1]{%
  \par\noindent
  \hangindent=2em
  \hangafter=1
  #1\par
  \medskip
}

Chapters \ref{chap:jtd_tismir}, \ref{chap:rhythm_rsos}, and \ref{chap:mp_network} of this thesis are respectively based on the following published journal articles:

\vspace{1.5em}

\bibitemhang{Cheston, H., Schlichting, J. L., Cross, I., and Harrison, P. M. C. (2024). Jazz Trio Database: Automated annotation of jazz piano trio recordings processed using audio source separation. \textit{Transactions of the International Society for Music Information Retrieval}, 7(1), 144--158. \url{https://doi.org/10.5334/tismir.186}}

\bibitemhang{Cheston, H., Schlichting, J. L., Cross, I., and Harrison, P. M. C. (2024). Rhythmic qualities of jazz improvisation predict performer identity and style in source-separated audio recordings. \textit{Royal Society Open Science}, 11(11), 240920. \url{https://doi.org/10.1098/rsos.240920}}

\bibitemhang{Cheston, H., Cross, I., and Harrison, P. M. C. (2024). Trade-offs in coordination strategies for duet jazz performances subject to network delay and jitter. \textit{Music Perception}, 42(1), 48--72. \url{https://doi.org/10.1525/mp.2024.42.1.48}}

\vspace{1.5em}

Chapter \ref{chap:xai_rsi} is based on the following preprint:

\vspace{1.5em}

\bibitemhang{Cheston, H., Bance, R., and Harrison, P. M. C. (2025). Deconstructing jazz piano style using machine learning. \textit{arXiv}. \url{https://doi.org/10.48550/arXiv.2504.05009}}

\vspace{1.5em}

Early versions of Chapters \ref{chap:jtd_tismir}, \ref{chap:rhythm_rsos}, and \ref{chap:mp_network} were posted as the following preprints:

\vspace{1.5em}

\bibitemhang{Cheston, H., Schlichting, J. L., Cross, I., and Harrison, P. M. C. (2024). Jazz Trio Database: Automated annotation of jazz piano trio recordings processed using audio source separation. \textit{PsyArXiv}. \url{https://doi.org/10.31234/osf.io/jyqp3}}

\bibitemhang{Cheston, H., Schlichting, J. L., Cross, I., and Harrison, P. M. C. (2024). Rhythmic qualities of jazz improvisation predict performer identity and style in source-separated audio recordings. \textit{PsyArXiv}. \url{https://doi.org/10.31234/osf.io/txy2f}}

\bibitemhang{Cheston, H., Cross, I., and Harrison, P. M. C. (2023). Trade-offs in coordination strategies for networked jazz performances. \textit{PsyArXiv}. \url{https://doi.org/10.31234/osf.io/z8c7w}}

\vspace{1.5em}

Early versions of work described in this thesis appeared in the following conference proceedings:

\vspace{1.5em}

\bibitemhang{Cheston, H., Cross, I., and Harrison, P. M. C. (2022). The effects of variable latency timings and jitter on networked musical proceedings. In \textit{Proceedings of the 15th International Conference of Students of Systematic Musicology}. Ghent, Belgium.}

\bibitemhang{Cheston, H., Cross, I., and Harrison, P. M. C. (2022). Measuring the effects of variable latency timings on networked jazz performances. In \textit{Proceedings of the Society for Education, Music, and Psychology Research 50th Anniversary Conference}. London, United Kingdom.}

\bibitemhang{Cheston, H., Cross, I., and Harrison, P. M. C. (2023). Modelling coordination strategies in improvised musical performances by skilled jazz groups. In \textit{Proceedings of the 16th International Conference of Students of Systematic Musicology}. Sheffield, United Kingdom.}

\bibitemhang{Cheston, H., Cross, I., and Harrison, P. M. C. (2023). Coordination strategies in networked jazz performances. In \textit{Proceedings of the 17th International Conference on Music Perception and Cognition}. Tokyo, Japan.}

\bibitemhang{Cheston, H., Cross, I., and Harrison, P. M. C. (2023). An automated pipeline for characterizing timing in jazz trios. In \textit{Proceedings of the 18th Digital Music Research Network Workshop}. London, United Kingdom.}

\bibitemhang{Cheston, H., Bance, R., and Harrison, P. M. C. (2024). Characterizing jazz improvisation style through explainable performer identification models. In \textit{Proceedings of the 19th Digital Music Research Network Workshop}. London, United Kingdom.}

\vspace{1.5em}

The following other papers were also produced during the doctoral study period:

\vspace{1.5em}

\bibitemhang{Cheston, H., Van Balen, J., and Durand, S. (2025). Automatic identification of samples in hip-hop music via multi-loss training and an artificial dataset. \textit{arXiv}. \url{https://doi.org/10.48550/arXiv.2502.06364}}
